Let's talk about Thurstone & Co.: An information-theoretical model for comparative judgments, and its statistical translation

### 2024-12-20 paper2.qmd
section:model; process:write

<!-- 1. Thurstone's Theory of CJ: discriminal process -->
In its most general form, Thurstone’s theory addresses pairwise comparisons where a single judge evaluates multiple stimuli [@Thurstone_1927b, pp. 267]. The theory posits that two key factors determine the dichotomous outcome of these comparisons: the discriminal process of each stimulus and their discriminal difference. The *discriminal process* captures the psychological impact each stimulus exerts on the judge or, more simply, his perception of the stimulus trait. The theory assumes that the discriminal process for any given stimulus forms a Normal distribution along the trait continuum [@Thurstone_1927b, pp. 266]. The mode (mean) of this distribution, known as the *modal discriminal process*, indicates the stimulus position on this continuum, while its dispersion, referred to as the *discriminal dispersion*, reflects variability in the perceived trait of the stimulus.

@@

<!-- 2. Example of discriminal processes -->
@fig-discriminal_process illustrates hypothetical discriminal processes along a quality trait continuum for two written texts. The figure indicates that the modal discriminal process for Text B is positioned further along the continuum than that of Text A $(T_{B} > T_{A})$, suggesting that Text B exhibits higher quality. Additionally, the figure highlights that Text B has a broader distribution compared to Text A, which arises from its larger discriminal dispersion $(\sigma_{B} > \sigma_{A})$.

@@

<!-- 2. But this is not the scenario we observe -->
In experimental practice, however, this process occurs in reverse. Researchers first observe the comparison outcome and then use the BTL model to infer the discriminal difference between the stimuli and their respective discriminal processes [@Thurstone_1927a, pp. 373]. Therefore, it is reasonable to assume that the outcome's ability to reflect the "true" differences between stimuli largely depends on the validity of the model's assumptions, particularly the assumption of equal dispersions. For instance, when researchers observe a sample of outcomes favoring Text B over Text A and correctly assume equal dispersions between the texts, the BTL model estimates a discriminal difference distribution that accurately represents the "true" discriminal difference of the texts. This scenario is illustrated in @fig-dispersion, where the model's discriminal difference distribution aligns with the "true" distribution, represented by the thick continuous line corresponding to $\sigma_{B}-\sigma_{A}=0$. The accuracy of these discriminal difference ensures reliable estimates for the texts’ discriminal processes [(citation needed?)]{style="color:red;"}.

@@

@@

<!-- 2. But this is not the scenario we observe -->
However, in practice, researchers approach this process in reverse. They begin by observing the sample of outcomes favoring Text B over Text A and then use the BTL model to estimate the discriminal difference and the discriminal processes of the stimuli. Given that the BTL model assumes independent discriminal processes across comparisons, if this assumption holds, then the model estimates a discriminal difference distribution that accurately reflects the "true" discriminal difference of the texts. This scenario is illustrated with @fig-correlation when the discriminal difference distribution of the model aligns with the "true" distribution, represented by the thick continuous line corresponding to $\rho=0$. Once more, the estimation accuracy of the discriminal difference ensures reliable estimates for the discriminal processes of the texts [(citation needed?)]{style="color:red;"}.

@@

This section introduces a theoretical model for Comparative Judgment (CJ) that builds upon Thurstone's theory. The model systematically accounts for the factors influencing judges during CJ experiments. These theoretical and practical insights are then translated into a statistical model.

@@

#### Figure 3

@@

#### Figure 4

@@

#### Figure 5

@@

#### Figure 6

@@

#### Figure 7

@@

#### Figure 8

@@

#### Figure 9

@@

#### Figure 10

@@

#### Figure 11

@@

#### Figure 12

@@

Considering the sampling process

@@

#### Figure 13

@@

#### Figure 14

@@

#### Figure 15

@@

>>>>

<!-- 1. Thurstone's Theory of CJ: discriminal process -->
In its most general form, Thurstone’s theory addresses pairwise comparisons where a single judge evaluates multiple stimuli [@Thurstone_1927b, pp. 267]. The theory posits that two key factors determine the dichotomous outcome of these comparisons: the discriminal process of each stimulus and their discriminal difference. The *discriminal process* captures the psychological impact each stimulus exerts on the judge or, more simply, his perception of the stimulus trait. The theory assumes that the DP for any given stimulus forms a Normal distribution along the trait continuum [@Thurstone_1927b, pp. 266]. The mode (mean) of this distribution, known as the *modal discriminal process*, indicates the stimulus position on this continuum, while its dispersion, referred to as the *discriminal dispersion*, reflects variability in the perceived trait of the stimulus.

@@

<!-- 2. Example of discriminal processes -->
@fig-discriminal_process illustrates hypothetical discriminal processes along a quality trait continuum for two written texts. The figure indicates that the mDP for Text B is positioned further along the continuum than that of Text A $(T_{B} > T_{A})$, suggesting that Text B exhibits higher quality. Additionally, the figure highlights that Text B has a broader distribution compared to Text A, which arises from its larger discriminal dispersion $(\sigma_{B} > \sigma_{A})$.

@@

<!-- 2. But this is not the scenario we observe -->
In experimental practice, however, this process occurs in reverse. Researchers first observe the comparison outcome and then use the BTL model to infer the discriminal difference between the stimuli and their respective discriminal processes [@Thurstone_1927a, pp. 373]. Therefore, it is reasonable to assume that the outcome's ability to reflect the "true" differences between stimuli largely depends on the validity of the model's assumptions [@Kohler_et_al_2019, pp. 150], particularly the assumption of equal dispersions. For instance, when researchers observe a sample of outcomes favoring Text B over Text A and correctly assume equal dispersions between the texts, the BTL model estimates a discriminal difference distribution that accurately represents the "true" discriminal difference of the texts. This scenario is illustrated in @fig-dispersion, where the model's discriminal difference distribution aligns with the "true" distribution, represented by the thick continuous line corresponding to $\sigma_{B}-\sigma_{A}=0$. The accuracy of these discriminal difference ensures reliable estimates for the texts’ discriminal processes [(citation needed?)]{style="color:red;"}.

@@

<!-- Statements about the real world from observed data require assumptions. This is true no matter what kind of data we have, and it is explicitly also true for statements based on data stemming from SRSg. The validity of our statements therefore critically hinges on the validity of those assumptions. If our statistical assumptions are wrong for the analysis at hand, the inferential statements will be incorrect no matter how sophisticated the statistical methods are. [@Kohler_et_al_2019, pp. 150] -->

@@

<!-- 2. But this is not the scenario we observe -->
Off course, in experimental practice, researchers approach this process in reverse. They begin by observing the sample of outcomes favoring Text B over Text A and then use the BTL model to estimate the discriminal difference and the discriminal processes of the stimuli. Given that the BTL model assumes independent discriminal processes across comparisons, if this assumption holds, then the model estimates a discriminal difference distribution that accurately reflects the "true" discriminal difference of the texts. This scenario is illustrated with @fig-correlation when the discriminal difference distribution of the model aligns with the "true" distribution, represented by the thick continuous line corresponding to $\rho=0$. Once more, the estimation accuracy of the discriminal difference ensures reliable estimates for the discriminal processes of the texts [(citation needed?)]{style="color:red;"}.

@@

This section presents a theoretical model for CJ that extends Thurstone's theory. It uses causal analysis [@Pearl_2009; @Pearl_et_al_2016; @Morgan_et_al_2014; @Hernan_et_al_2020] and, in particular, directed acyclic graphs (DAGs) [@Gross_et_al_2018; @Neal_2020], to articulate the core theoretical principles of CJ theory. The model also incorporates several practical factors that influence judges in CJ experiments, such as the selection of judges, stimuli, and the number of comparisons. In addition, the study uses Bayesian statistical methods [@McElreath_2020] to translate these theoretical and practical elements into a statistical model that facilitates the analysis of pairwise comparison data.

@@

#### Figure

#### 3

@@

#### 4

@@

#### 5

@@

#### 6

@@

#### Figure

#### 7

@@

#### 8

@@

#### 9

@@

#### 10

@@

#### 11

@@

#### 12

@@

Considering the sampling and comparison mechanisms

@@

#### Figure

#### 13

@@

#### 14

@@

#### 15

@@

<!-- The condition of case selection to be independent from Y is always met in experiments because the Y is not realized before the experiment starts. [@Kohler_et_al_2019, pp. 160] -->
